### 2020-02-14 doc/icfp20/counterpoint.tex
tweaks

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Figure~\ref{fig:fs} is an example of first species counterpoint.
The lower line is the cantus firmus, which we take from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

The reason behind our three-part representation of music is that
different parts are subject to different rules.
Let us go through the

@@

The beginning of the music should express perfection.
Therefore, the first interval must be a unison, a 5th, or an 8th.
In our formalization, we implement this rule as a function
\texttt{checkBeginning}, which reports an error \texttt{not158 i}
when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we wrap this error around
the \texttt{just} constructor for option values.

@@

The middle bars of the music should maintain harmonical
consonance and independence of melodic lines.
This means, all constituent intervals must be chosen from the 3rd,
5th, 6th, and 8th.
We encode this rule as the \texttt{checkIntervals} function,
which returns a list of errors correponding to the occurrences of
dissonant intervals and unisons.

@@

checkMotion : List PitchInterval \(\rightarrow\) List MotionError
checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs  
\end{alltt}

@@

The ending of the music should express relaxation.
This means, the last interval must be chosen from the unison and
the 8th, which are both considered very stable.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
We encode these rules as the \texttt{checkEnding} function;
note that it returns the \texttt{tooShort} error when the music does
not have enough number of bars to form a valid ending.

@@

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Roughly speaking, strong beats in middle bars are constrained by the
same rules as in first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.
Weak beats, on the other hand, are constrained in a looser manner.
First, they are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
Second, they are allowed to be a unison if they are left by step in the
opposite direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).

>>>>

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint.
The lower line is the cantus firmus, which we excerpt from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

The reason behind our three-part representation of music is that
different parts are subject to different rules, as we detail below.

@@

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the unison, the 5th, and the 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into an option
value by wrapping it around the \texttt{just} constructor.

@@

The middle bars of the music should maintain harmonical consonance
and independence of melodic lines.
As we saw before, consonant intervals include the unison, 3rd, 5th,
6th, and 8th, and among these, the unison is clearly an obstacle to
distinguish between the two lines of music.
Therefore, the middle bars must consist of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors correponding to the occurrences of dissonant
intervals and unisons.

@@

checkMotion : List PitchInterval \(\rightarrow\) List MotionError
checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs
-- where pairs returns a list of all adjacent pairs in a given list
\end{alltt}

@@

The ending of the music should express relaxation.
Among different intervals, the unison and the 8th are considered
most stable, hence the last interval of the music must be either of
these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
We encode these rules as the \texttt{checkEnding} function.
Note that it returns the \texttt{tooShort} error when the music does
not have enough number of bars to form a valid ending.

@@

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Roughly speaking, strong beats in middle bars are constrained by the
same rules as in first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.

Weak beats, on the other hand, are constrained in a looser manner.
First, they are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
Second, they are allowed to be a unison if they are left by step in the
opposite direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).

### 2020-02-14 doc/icfp20/music.tex
tweaks

\paragraph{Pitches}
In music, a \emph{pitch} is a value denoting how high or low a sound is.
Pitches can either be represented either as an absolute value or as a
relative value paired with the octave the sound belongs to.
In our implementation, the two representations of pitches are defined
as \texttt{Pitch} and \texttt{PitchOctave}, respectively, with the proof
that converting back and forth between the two is an identity.

@@

\paragraph{Intervals}
\emph{Intervals} are yet another key concept in music.
An interval denotes the difference in pitch between two notes.
There are 13 kinds of interval within an octave, and these intervals
may be classified from several different perspectives: (i) major or minor;
(ii) consonant or dissonant; and (iii) perfect or imperfect.
For convenience, we let the names of intervals carry the information
about the first and third classifications.

>>>>

\paragraph{Pitches}
In music, a \emph{pitch} is a value denoting how high or low a sound is.
Pitches can be represented either as an absolute value or as a relative
value paired with the octave the sound belongs to.
In our implementation, the two representations of pitches are defined
as \texttt{Pitch} and \texttt{PitchOctave}, respectively, accompagnied
by a proof that converting back and forth between them is an identity.

@@

\paragraph{Intervals}
\emph{Intervals} are yet another key concept in music.
An interval denotes the difference in pitch between two notes.
There are 13 kinds of interval within an octave, and these intervals
may be classified from several different perspectives: (i) major or minor;
(ii) consonant or dissonant; and (iii) perfect or imperfect.
For convenience, we let the names of intervals carry the information
about the first and third classifications.
As a convention, we refer to \texttt{min2} and \texttt{maj2}  uniformly
as ``2nd'', and similarly for other intervals.
We also call \texttt{per1} the \emph{unison} in the rest of this section.
